---
title: "Explore Freeze-dried samples hellinger and CLR beta diversity"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    fig_caption: yes  
    toc: yes
    extra_dependencies: ["flafter"]
---


```{r setup, echo=F,  message=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.path='figures/CLRHell/', cache = TRUE)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggdendro))
suppressPackageStartupMessages(library(harrypotter))
suppressPackageStartupMessages(library(ape))
```

```{r constants}
```

```{r read-data, echo= FALSE, message=FALSE}
asvs <- read_tsv("data/asv_summed.tsv", col_types = 'ccidd')
metadata <- read_csv("data/FDvsFrozenMetadata.csv", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = ""),
  SampleOrder = col_number()
)) %>% suppressWarnings()
taxonomy <- read_tsv("data/taxonomy.tsv", col_types =  cols(
  asv = col_character(),
  kingdom = col_character(),
  phylum = col_character(),
  class = col_character(),
  order = col_character(),
  family = col_character(),
  genus = col_character(),
  species = col_character()
))

# Rename Sample number
metadata  <- metadata %>% dplyr::rename(OldSampleNumber = SampleNumber, SampleNumber = NewSampleNumber )
# make a character of the sample number
metadata <- metadata %>% mutate(Asample = paste("A", sample, sep= "_"))

# filter out the 40 samples freeze dried and their frozen comparison
asvs <- asvs %>% inner_join(metadata %>% dplyr::select(sample), by ="sample") 


asvs <- asvs %>% left_join(metadata %>% select(sample,  Asample))

# get palette
pal <- hp(n = 8, house = "ronweasley2")
pal <- pal[c(6,4)]

```


# Introducrtion

I here use Hellinger transformed counts which puts more weight to common ASVs) and clr-transformed approach *for more information on clr, see Gloor et al. 2017 and Gloor's github tutorial* ( https://github.com/ggloor/CoDa_microbiome_tutorial/wiki/Part-1:-Exploratory-Compositional-PCA-biplot ) to explore beta diversity of freeze-dried and frozen samples.


 
# Hellinger

```{r pca-hell, echo=FALSE, fig.width=12, fig.height = 12, fig.cap = '**Fig 1. Screeplot and dendrogram for pca on hellinger transformed counts**. Samples group with Population origin.', message=FALSE, warning=FALSE}
hellingers <- asvs %>% select(asv, sample, count) %>%
  pivot_wider(names_from = 'asv', values_from = 'count', values_fill = 0) %>%
  as.data.frame() %>% tibble::column_to_rownames('sample') %>%
  decostand(method = 'hellinger') %>%
  data.frame() %>%
  tibble::rownames_to_column('sample') %>%
  pivot_longer(names_to = 'asv', values_to = 'hellinger', 2:ncol(.)) %>%
  mutate(asv = str_remove(asv, '^X'))

hell <- hellingers %>% 
  spread(asv, hellinger, fill = 0) %>%
  tibble::column_to_rownames('sample') %>% 
  as.matrix()
hell %>%  prcomp() -> pca

# Calculate the variance explained by PC1 and PC2
d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="")
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="")
```



```{r calc-hellinger,  fig.cap='**Figure 1. PCA Hellinger Transformed counts.** ', fig.height=6, fig.width=12, message=FALSE, warning=FALSE, echo=FALSE}
#Add metadata to pca df
df<-data.frame(pca$x)
df$sample<-rownames(df)
df1 <- df %>% left_join(metadata)
#df1[1:5,c(1:3,109:140)]

#PCA representation

# plot
PCAhell1 <- 
  ggplot(df1, aes(x=PC1,y=PC2, group = SampleNumber)) + 
  geom_point( aes(col=Treatment), size = 4) +  
  geom_line( color="black", aes(alpha = 1)) +
    scale_colour_manual( values = pal) +
                      ylab(PC2)+ xlab(PC1) + 
                                        theme(axis.title=element_text(size=18,face="bold"),
                                        axis.text = element_text(size=14),
                                        panel.background=element_rect(fill='white', colour='black'),
                                        strip.background=element_rect(fill='white', colour='white'),
                                        panel.grid = element_blank(),
                                        legend.key=element_blank(),
                                        legend.position = c(x= 0.1, y = 0.85),
                                        legend.title = element_blank(),
                                        legend.text = element_text( size=14, face="bold")) +
  guides(alpha =FALSE, fill =FALSE)

PCAhell1
```



#### Permanovas hellinger
```{r permanova-adonis-calc-hellinger, message=FALSE, warning=FALSE, include=FALSE}
set.seed(1000)
metadf <- metadata %>% arrange(sample) %>% tibble::column_to_rownames('sample') %>% data.frame()
hell <- asvs %>% select(asv, sample, count) %>% arrange(sample) %>%
  pivot_wider(names_from = 'asv', values_from = 'count', values_fill = 0) %>%
  as.data.frame() %>% tibble::column_to_rownames('sample') %>%
  decostand(method = 'hellinger') %>%
  data.frame() %>% dist(method = "euclidian")

# adonis2
perm <- how(nperm = 9999)
permanova <- adonis2(hell ~ Plate_No, data = metadf, permutations = perm,  by = "margin")
print(as.data.frame(permanova))

betadisp <- betadisper(hell, metadf$Plate_No, 
                                     type = "centroid", bias.adjust = FALSE, sqrt.dist = FALSE, add = FALSE)
#"boxplot"(betadisp, ylab = "Distance to centroid", xlab = "Treatment") # Doesn't look OK... 
anova(betadisp)

permanova <- adonis2(hell ~ Treatment + SampleNumber, data = metadf, permutations = perm,  by = "margin")
print(as.data.frame(permanova))
betadisp <- betadisper(hell, metadf$Plate_No, 
                                     type = "centroid", bias.adjust = FALSE, sqrt.dist = FALSE, add = FALSE)
#"boxplot"(betadisp, ylab = "Distance to centroid", xlab = "Treatment") # Doesn't look OK... 
anova(betadisp)

setBlocks(perm) <- with(metadf, Plate_No)
permanova <- adonis2(hell ~ Treatment + SampleNumber, data = metadf, permutations = perm,  by = "margin")
print(as.data.frame(permanova))
```





# CLR

$~$


I've calcutad clr using the compositions-package and adding the psuedocount **0.65** to each ASV/sample count.

$~$


```{r clr, include=FALSE}
# make a matrix with row sample and column ASV with counts
m1 <- asvs  %>% select(asv,Asample) %>%
  complete(asv, Asample) %>% left_join(asvs %>% select(asv, count, Asample)) %>%
  #left_join(metadata %>% select(Asample, NbReads))%>% 
  mutate( count = replace_na(count, 0),
          count = count +0.65) %>% # add psuedo count of 1/ library size
  pivot_wider(names_from = 'asv', values_from='count') %>% 
  tibble::column_to_rownames('Asample') %>% as.matrix()
#m1[1:5,1:5]#row=samples

### clr-transfornmation of the counts ###
#pseudocount <- 0.65 #add a pseudo count for zeroes
clr <- t(apply(t(m1), 2, compositions::clr))
#clr[1:5,1:5]#row=samples
#dist(clr)#Aitchison distance
```



```{r PCAclr, message=FALSE, warning=FALSE, include=FALSE}
### Perform the PCA ###
pca <- prcomp(clr)
#plot(pca$x[,1],pca$x[,2])

# Calculate the variance explained by PC1 and PC2
d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(sum(pca$sdev[1]^2)/d.mvar, 3)*100,"%",")",sep="")
PC2 <- paste("PC2 ","(",round(sum(pca$sdev[2]^2)/d.mvar, 3)*100,"%",")",sep="")
```




```{r pc1pc2, echo=FALSE, fig.width=12, fig.height = 6, fig.cap = '**Fig 2. PCA on clr transformed counts.**', message=FALSE, warning=FALSE}
#Add metadata to pca df
df<-data.frame(pca$x)
df$Asample<-rownames(df)
df1 <- df %>% left_join(metadata)
#df1[1:5,c(1:3,109:140)]

#PCA representation

# plot by population
PCACLR1<-ggplot(df1, aes(x=PC1,y=PC2, group = SampleNumber)) + 
  geom_point( aes(col=Treatment), size = 4) +  
  geom_line( color="black", aes(alpha = 1)) +
    scale_colour_manual( values = pal) +
                      ylab(PC2)+ xlab(PC1) + 
                                        theme(axis.title=element_text(size=18,face="bold"),
                                        axis.text = element_text(size=14),
                                        panel.background=element_rect(fill='white', colour='black'),
                                        strip.background=element_rect(fill='white', colour='white'),
                                        panel.grid = element_blank(),
                                        legend.key=element_blank(),
                                        legend.position = c(x= 0.1, y = 0.9),
                                        legend.title = element_blank(),
                                        legend.text = element_text( size=14, face="bold")) +
  guides(alpha =FALSE, fill =FALSE)

PCACLR1    
```



#### Permanovas Aitchison/Euclidian distance of clr transformation
```{r clrperma, message=FALSE, warning=FALSE, include=FALSE}

clrdist<-dist(clr) #Aitchison/Euclidian distance 
metadf <- metadata %>% tibble::column_to_rownames('Asample')

set.seed(1000)

perm <- how(nperm = 9999)
permanova <- adonis2(clrdist ~ Plate_No, data = metadf, permutations = perm,  by = "margin")
print(as.data.frame(permanova))

betadisp <- betadisper(clrdist, metadf$Plate_No, type = "centroid", bias.adjust = FALSE, sqrt.dist = FALSE, add = FALSE)
#"boxplot"(betadisp, ylab = "Distance to centroid", xlab = "Treatment") # Doesn't look OK... 
anova(betadisp)

permanova <- adonis2(clrdist ~ Treatment + SampleNumber, data = metadf, permutations = perm,  by = "margin")
print(as.data.frame(permanova))

# with plate no as strata arg
setBlocks(perm) <- with(metadf, Plate_No)
permanova <- adonis2(clrdist ~ Treatment + SampleNumber, data = metadf, permutations = perm,  by = "margin")
print(as.data.frame(permanova))

betadisp <- betadisper(clrdist, metadf$Treatment, type = "centroid", bias.adjust = FALSE, sqrt.dist = FALSE, add = FALSE)
#"boxplot"(betadisp, ylab = "Distance to centroid", xlab = "Treatment") # Doesn't look OK... 
anova(betadisp)

```


 
# combine plots

```{r combinedPCA, echo=FALSE, fig.cap='**Figure 14.Combined PCA Hellinger and CLR.**', fig.height=5, fig.width=12, message=FALSE, warnings=FALSE}
ggarrange(PCAhell1 + ggtitle("A) Hellinger"),NULL, PCACLR1+ ggtitle("B) CLR"), widths = c(1, 0.05, 1), nrow= 1, ncol =3, common.legend = TRUE)
```
